---
title: CRAN memory error
author: Brian J. Knaus
date: '2017-06-12'
slug: cran-memory-error
categories: [R]
tags: [R, Rcpp, ASAN, UBSAN]
description: ''
draft: no
---


I have a package on CRAN where the CRAN tests are reporting a memory error.
In order to use the address sanitizer (ASAN) we need to have a version of R that has been compiled specially to make use of it.
In a previous post I described how we can use [Docker to run rocker images](../running-r-in-docker/) so we do not have to change our system R.
I've also validated the [ASAN is working](../validating-asan/).
We're alost ready to diagnose our issue.
First we'll need to install some software.
We'll also need to handle an R library issue.


## Installing software


In orderer to build the vignettes we'll need a few dependencies from outside R.
We can install these in our Docker instance in a manner similar to a typical system.


```{R, eval = FALSE}
apt-get update
apt-get install libssh2-1-dev libssl-dev pandoc qpdf
```


We'll also need to install some package dependencies.
We can do this from the shell as follows.


```{R, eval = FALSE}
R -e 'install.packages(c("ape", "dplyr", "knitr", "poppr", "Rcpp", "memuse", "pinfsc50", "rmarkdown", "testthat", "tidyr", "vegan", "viridisLite"), dependencies = TRUE, lib = "/usr/local/lib/R/site-library")'
```


This actually take a bit due to the number of packages that are needed.


## Build and check the package


You should now be able to build and check your package.


```{R, eval = FALSE}
cd Rsource
R CMD build vcfR/
R CMD check --as-cran vcfR_1.4.0.9000.tar.gz
```



### A package finding error

I generated a curious error that took me a bit to track down.
When trying to build my package I generate the below error.

```{bash, eval = FALSE}
Error in loadNamespace(j <- i[[1L]], c(lib.loc, .libPaths()), versionCheck = vI[[j]]) : 
  there is no package called ‘lattice’
ERROR: lazy loading failed for package ‘vcfR’
```


I say this is curious because when I enter R and try to load this package interactively it seems to work fine.
For some reason it is not found during the build invocation.
I find I can circumvent this issue by moving packages to a specific location.
On my system packages may be located at several locations.


```{bash, eval = FALSE}
.libPaths()
"/usr/local/lib/R/site-library"
"/usr/local/lib/R/library"
"/usr/lib/R/library" 
```


I can remove packages and install them to a specific location like this.


```{bash, eval = FALSE}
find.package('lattice')
remove.packages('lattice', lib = "/usr/lib/R/library")
find.package('lattice')
install.packages('lattice', lib = "/usr/local/lib/R/site-library")
```


I find that this issue affects more than one package, so I tend to move everything I can.
Note that you don't want to move the base packages, so I omit them.


```{bash, eval = FALSE}
# base packages, and should not be updated
myBase <- c('base', 'compiler', 'datasets', 'graphics', 'grDevices', 'grid', 'methods', 'parallel', 'splines', 'stats', 'stats4', 'tcltk', 'tools', 'utils')

# First directory
miss.pkgs <- list.files("/usr/lib/R/library")
miss.pkgs <- grep(paste(myBase, collapse = "|"), miss.pkgs, value = TRUE, invert = TRUE)
remove.packages(miss.pkgs, lib = "/usr/lib/R/library")
install.packages(miss.pkgs, lib = "/usr/local/lib/R/library")

# Second directory
miss.pkgs <- list.files("/usr/local/lib/R/site-library")
miss.pkgs <- grep(paste(myBase, collapse = "|"), miss.pkgs, value = TRUE, invert = TRUE)
remove.packages(miss.pkgs, lib = "/usr/local/lib/R/site-library")
install.packages(miss.pkgs, lib = "/usr/local/lib/R/library")

```


Once we have the required dependencies we can build and check our package.


## Build and test package


If you have not mounted a volume containing the package you could download it from CRAN.


```{r, eval = FALSE}
mkdir RSource
cd RSource
wget https://cran.r-project.org/src/contrib/vcfR_1.4.0.tar.gz
tar -xvzf vcfR_1.4.0.tar.gz
cd ..
```


Build and test the package as follows.


```{r, eval = FALSE}
Rdevel CMD build vcfR/
Rdevel CMD check --as-cran vcfR_1.4.0.9000.tar.gz 
```


```{}
# Rdevel CMD INSTALL vcfR_1.4.0.9000.tar.gz 

# RD CMD build /RSource/vcfR
# RD CMD check --as-cran vcfR_1.4.0.9000.tar.gz
# RD CMD check --no-build-vignettes vcfR_1.4.0.9000.tar.gz
```



I use `testthat` for unit testing.
One of my tests failed.
I can find that output in my tests directory.


```{bash, eval = FALSE}
less vcfR.Rcheck/tests/testthat.Rout.fail 
```


This should allow us to reproduce any error reported to us from CRAN.
If we can reproduce the error we should be able to fix it.


If there is a question about how the compiler was called we can check in the log files.
In the file `00install.out` we can validate that the compilation included the correct options.


```{bash, eval = FALSE}
-fsanitize=address -fno-omit-frame-pointer
```


In order to isolate the problem we can install the built package and try to execute the script that threw the error.


```{r, eval = FALSE}
Rdevel CMD INSTALL vcfR_1.4.0.tar.gz
Rscript RSource/vcfR/tests/testthat/test_3_extract_gt.R
```



```{bash, eval = FALSE}
root@f91ccb9ab8cf:/# Rscriptdevel RSource/vcfR/tests/testthat/test_3_extract_gt.R 

   *****       ***   vcfR   ***       *****
   This is vcfR 1.4.0
     browseVignettes('vcfR') # Documentation
     citation('vcfR') # Citation
   *****       *****      *****       *****

=================================================================
==6066==ERROR: AddressSanitizer: stack-buffer-overflow on address 0x7ffe1d3c3f83 at pc 0x7f22cb00bec0 bp 0x7ffe1d3c3650 sp 0x7ffe1d3c3648
READ of size 1 at 0x7ffe1d3c3f83 thread T0
    #0 0x7f22cb00bebf in gt2alleles(Rcpp::String, std::vector<std::__cxx11::basic_string<char, std::char_traits<char>, std::allocator<char> >, std::allocator<std::__cxx11::basic_string<char, std::char_traits<char>, std::allocator<char> > > >) (/usr/local/lib/R/site-library/vcfR/libs/vcfR.so+0x30bebf)
    #1 0x7f22cb014ea1 in extract_GT_to_CM2(Rcpp::Matrix<16, Rcpp::PreserveStorage>, Rcpp::Matrix<16, Rcpp::PreserveStorage>, std::__cxx11::basic_string<char, std::char_traits<char>, std::allocator<char> >, int, int, int) (/usr/local/lib/R/site-library/vcfR/libs/vcfR.so+0x314ea1)
    #2 0x7f22cafde0dd in vcfR_extract_GT_to_CM2 (/usr/local/lib/R/site-library/vcfR/libs/vcfR.so+0x2de0dd)
    #3 0x7f22da6f126b in do_dotcall (/usr/local/lib/R/lib/libR.so+0xa8226b)
    #4 0x7f22da7ed244 in bcEval (/usr/local/lib/R/lib/libR.so+0xb7e244)
    #5 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #6 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #7 0x7f22da82f700 in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc0700)
    #8 0x7f22da843c56 in do_set (/usr/local/lib/R/lib/libR.so+0xbd4c56)
    #9 0x7f22da82fc98 in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc0c98)
    #10 0x7f22da83d36d in do_begin (/usr/local/lib/R/lib/libR.so+0xbce36d)
    #11 0x7f22da82fc98 in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc0c98)
    #12 0x7f22da847beb in do_eval (/usr/local/lib/R/lib/libR.so+0xbd8beb)
    #13 0x7f22da7ed244 in bcEval (/usr/local/lib/R/lib/libR.so+0xb7e244)
    #14 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #15 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #16 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #17 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #18 0x7f22da8323b4 in forcePromise (/usr/local/lib/R/lib/libR.so+0xbc33b4)
    #19 0x7f22da833643 in getvar (/usr/local/lib/R/lib/libR.so+0xbc4643)
    #20 0x7f22da7fe132 in bcEval (/usr/local/lib/R/lib/libR.so+0xb8f132)
    #21 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #22 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #23 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #24 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #25 0x7f22da8323b4 in forcePromise (/usr/local/lib/R/lib/libR.so+0xbc33b4)
    #26 0x7f22da833643 in getvar (/usr/local/lib/R/lib/libR.so+0xbc4643)
    #27 0x7f22da7fe132 in bcEval (/usr/local/lib/R/lib/libR.so+0xb8f132)
    #28 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #29 0x7f22da8323b4 in forcePromise (/usr/local/lib/R/lib/libR.so+0xbc33b4)
    #30 0x7f22da833643 in getvar (/usr/local/lib/R/lib/libR.so+0xbc4643)
    #31 0x7f22da7fe132 in bcEval (/usr/local/lib/R/lib/libR.so+0xb8f132)
    #32 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #33 0x7f22da8323b4 in forcePromise (/usr/local/lib/R/lib/libR.so+0xbc33b4)
    #34 0x7f22da833643 in getvar (/usr/local/lib/R/lib/libR.so+0xbc4643)
    #35 0x7f22da7fe132 in bcEval (/usr/local/lib/R/lib/libR.so+0xb8f132)
    #36 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #37 0x7f22da8323b4 in forcePromise (/usr/local/lib/R/lib/libR.so+0xbc33b4)
    #38 0x7f22da833643 in getvar (/usr/local/lib/R/lib/libR.so+0xbc4643)
    #39 0x7f22da7fe132 in bcEval (/usr/local/lib/R/lib/libR.so+0xb8f132)
    #40 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #41 0x7f22da8323b4 in forcePromise (/usr/local/lib/R/lib/libR.so+0xbc33b4)
    #42 0x7f22da833643 in getvar (/usr/local/lib/R/lib/libR.so+0xbc4643)
    #43 0x7f22da7fe132 in bcEval (/usr/local/lib/R/lib/libR.so+0xb8f132)
    #44 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #45 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #46 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #47 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #48 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #49 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #50 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #51 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #52 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #53 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #54 0x7f22da8323b4 in forcePromise (/usr/local/lib/R/lib/libR.so+0xbc33b4)
    #55 0x7f22da833643 in getvar (/usr/local/lib/R/lib/libR.so+0xbc4643)
    #56 0x7f22da7fe132 in bcEval (/usr/local/lib/R/lib/libR.so+0xb8f132)
    #57 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #58 0x7f22da8323b4 in forcePromise (/usr/local/lib/R/lib/libR.so+0xbc33b4)
    #59 0x7f22da833643 in getvar (/usr/local/lib/R/lib/libR.so+0xbc4643)
    #60 0x7f22da7fe132 in bcEval (/usr/local/lib/R/lib/libR.so+0xb8f132)
    #61 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #62 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #63 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #64 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #65 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #66 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #67 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #68 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #69 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #70 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #71 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #72 0x7f22da8009b7 in bcEval (/usr/local/lib/R/lib/libR.so+0xb919b7)
    #73 0x7f22da82f22f in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc022f)
    #74 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #75 0x7f22da82f700 in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc0700)
    #76 0x7f22da83d36d in do_begin (/usr/local/lib/R/lib/libR.so+0xbce36d)
    #77 0x7f22da82fc98 in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc0c98)
    #78 0x7f22da839d41 in R_execClosure (/usr/local/lib/R/lib/libR.so+0xbcad41)
    #79 0x7f22da82f700 in Rf_eval (/usr/local/lib/R/lib/libR.so+0xbc0700)
    #80 0x7f22da8d651d in Rf_ReplIteration (/usr/local/lib/R/lib/libR.so+0xc6751d)
    #81 0x7f22da8d7218 in R_ReplConsole (/usr/local/lib/R/lib/libR.so+0xc68218)
    #82 0x7f22da8d7327 in run_Rmainloop (/usr/local/lib/R/lib/libR.so+0xc68327)
    #83 0x55bbbebad9f9 in main (/usr/local/lib/R/bin/exec/R+0x9f9)
    #84 0x7f22d89ae2b0 in __libc_start_main (/usr/lib/x86_64-linux-gnu/libc.so.6+0x202b0)
    #85 0x55bbbebada49 in _start (/usr/local/lib/R/bin/exec/R+0xa49)

Address 0x7ffe1d3c3f83 is located in stack of thread T0 at offset 579 in frame
    #0 0x7f22cb0082af in gt2alleles(Rcpp::String, std::vector<std::__cxx11::basic_string<char, std::char_traits<char>, std::allocator<char> >, std::allocator<std::__cxx11::basic_string<char, std::char_traits<char>, std::allocator<char> > > >) (/usr/local/lib/R/site-library/vcfR/libs/vcfR.so+0x3082af)

  This frame has 9 object(s):
    [32, 36) 'unphased_as_na'
    [96, 100) 'allele_number'
    [160, 168) '__size'
    [224, 232) '__osize'
    [288, 312) 'gt_vector'
    [352, 376) 'delim_vector'
    [416, 448) 'sep'
    [480, 512) 'na_allele'
    [544, 576) 'gt2' <== Memory access at offset 579 overflows this variable
HINT: this may be a false positive if your program uses some custom stack unwind mechanism or swapcontext
      (longjmp and C++ exceptions *are* supported)
SUMMARY: AddressSanitizer: stack-buffer-overflow (/usr/local/lib/R/site-library/vcfR/libs/vcfR.so+0x30bebf) in gt2alleles(Rcpp::String, std::vector<std::__cxx11::basic_string<char, std::char_traits<char>, std::allocator<char> >, std::allocator<std::__cxx11::basic_string<char, std::char_traits<char>, std::allocator<char> > > >)
Shadow bytes around the buggy address:
  0x100043a707a0: 00 00 00 00 00 00 00 00 f1 f1 f1 f1 04 f4 f4 f4
  0x100043a707b0: f2 f2 f2 f2 04 f4 f4 f4 f2 f2 f2 f2 00 f4 f4 f4
  0x100043a707c0: f2 f2 f2 f2 00 f4 f4 f4 f2 f2 f2 f2 00 00 00 f4
  0x100043a707d0: f2 f2 f2 f2 00 00 00 f4 f2 f2 f2 f2 00 00 00 00
  0x100043a707e0: f2 f2 f2 f2 00 00 00 00 f2 f2 f2 f2 00 00 00 00
=>0x100043a707f0:[f3]f3 f3 f3 00 00 00 00 00 00 00 00 00 00 00 00
  0x100043a70800: 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00
  0x100043a70810: 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00
  0x100043a70820: 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00
  0x100043a70830: 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00
  0x100043a70840: 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00 00
Shadow byte legend (one shadow byte represents 8 application bytes):
  Addressable:           00
  Partially addressable: 01 02 03 04 05 06 07 
  Heap left redzone:       fa
  Heap right redzone:      fb
  Freed heap region:       fd
  Stack left redzone:      f1
  Stack mid redzone:       f2
  Stack right redzone:     f3
  Stack partial redzone:   f4
  Stack after return:      f5
  Stack use after scope:   f8
  Global redzone:          f9
  Global init order:       f6
  Poisoned by user:        f7
  Container overflow:      fc
  Array cookie:            ac
  Intra object redzone:    bb
  ASan internal:           fe
  Left alloca redzone:     ca
  Right alloca redzone:    cb
==6066==ABORTING
root@f91ccb9ab8cf:/# 
```


This reproduces the error we're looking for.
(We knew which file it was from the report on CRAN.)
We should now be able to isolate the offending commands in this file.
This tells us at least two important issues with our code.
First, the problem is in the function `gt2alleles()`.
Second, the issue is with the object `gt2`.
However, this does not pinpoint the problem.
In order to accomplish that, we'll need to use a debugger.


```{bash, eval = FALSE}
shell> Rdevel -d gdb
(gdb) run
R> library(vcfR)
<Crtl + c>
(gdb) b gt2alleles
(gdb) c
```


We can now copy and paste our offending code into the terminal.


```{bash, eval = FALSE}
R> data(vcfR_example)
R> chrom <- create.chromR(name="Chrom", vcf=vcf, seq=dna, ann=gff, verbose=FALSE)
R> chrom <- masker(chrom, min_DP = 1e3, max_DP = 2e3)

R> # This is where our problem is.
R> gt <- extract.gt(chrom, element="GT", return.alleles = TRUE)

(gdb) s
(gdb) n

info args


print gt
print allele_vector


(gdb) fr v

(gdb) finish
(gdb) quit


```



```{bash, eval = FALSE}
quit
```




We can enter the R environment and copy and paste the lines of our test into the console until we generate our error.


```{bash, eval = FALSE}
Rdevel
#Rscriptdevel -e ''
#Rscriptdevel -e 'library(vcfR); '

```


Once we have found the offending line we'll want to use a debugger to step isolate the code at issue.
Kevin Ushey has put together a nice post called [Debugging with LLDB](http://kevinushey.github.io/blog/2015/04/13/debugging-with-lldb/) which addresses the use of lldb which is very similar to gdb.




