---
title: R 3.4.1 personal library location
author: Brian J. Knaus
date: '2017-07-06'
slug: r-3-4-1-personal-library-location
categories: [R]
tags: [R]
description: 'R 3.4.1 personal library location error'
---


R 3.4.1 'Single Candle' was released on June 30, 2017.
Our group is preparing to provide an R workshop at our societal meeting this summer, so I'm tryingot keep up to date with the R releases.
Installation on my work Ubuntu 16.04 LTS machine appeared to go smoothly.
However, upon trying to install packages I received the following error.


```{r, eval = FALSE}
Warning in install.packages(update[instlib == l, "Package"], l, contriburl =
contriburl,  :
  'lib = "/usr/lib/R/site-library"' is not writable
Would you like to use a personal library instead?  (y/n) y
Would you like to create a personal library
NA
to install packages into?  (y/n) y
Error in install.packages(update[instlib == l, "Package"], l, contriburl =
contriburl,  :
  unable to create ‘NA’
```


This seemed odd to me because I am used to R creating a personal library as a child of my home directory.
Here it is attempting to use 'NA", which fails.
With a little Googling I found the issue posted [here on stack overflow](https://stackoverflow.com/q/44861967), where Dirk Eddelbuettel has replied, and [on Debian Bug report logs](https://bugs.debian.org/cgi-bin/bugreport.cgi?bug=866768), where Dirk has also replied.
I also found [this post on stackoverflow](https://stackoverflow.com/a/29970899) useful, which is also by Dirk, because it explains the recommendation for where packages should be installed on a Unix system.
Here I explain how I used information from these links to start installing packages on my system.


## My system


First off its important to state that the system I'm discussing today is a work system on a work network, as opposed to a personal system.
That means I need to avoid interrupting any network settings or else I'll have to contact IT to get them to help me fix it.
Fortunately, our netowrk is a Windows network, so I'm really more of an exception which may give me more flexibility than others.


## My .libPaths()


As suggested by Dirk I checked the permissions of my `/usr/local/lib/R/site-library`.


```{r, eval=FALSE}
$ ls -l /usr/local/lib/R/
total 4
drwxrwsr-x 2 root staff 4096 Aug  5  2016 site-library
```


Its owned by `root` and group memebership belongs to `staff`.
I can check and see that I'm not a part of `staff`, which explains the error.


```{r, eval=FALSE}
$ id
uid=1142428492(knausb) gid=1142424065(domain^users) groups=1142424065(domain^users),126(vboxusers),1142425304(grunwald-lab),1142428469(linux-users),1142429799(certsvc_dcom_access),1142429897(linux^admins)
```


Because I'm not actually the system administrator, I do not know a lot about the existing groups.
I found [this post](https://stackoverflow.com/a/2835475) which explains how I can query for the members of groups on my system.
After creating the script I use the following to determine that there are no members of this group.


```{r, eval=FALSE}
perl -T list_group_members.pl staff
```


This leads me to believe I can change it's group ownership with no consequences.


```{r, eval=FALSE}
$ sudo chgrp domain^users /usr/local/lib/R/site-library/
$ ls -l /usr/local/lib/R/
total 4
drwxrwsr-x 103 root domain^users 4096 Jul  6 10:16 site-library
```


Now I, and presumedly other users on this system if there are any, can install packages into `/usr/local/lib/R/site-library/`.
Which means I have some package installations to work on.

