---
title: Hello blogdown
author: Brian J. Knaus
date: '2017-05-26'
slug: hello-blogdown
categories: [R]
tags: [blogdown]
description: 'First post'
thumbnail: ''
---

This is my first blog. Hello blogdown!


I built this site using the directions I found at [this link](https://proquestionasker.github.io/blog/Making_Site/) provided by Amber Thomas.
Note that the scripts 'setup.sh' and 'deploy.sh' work on your GitHub repository names 'username.github.io'.
Even if you think you're building it somewhere else, this script will clobber the contents of that repository.
Also, try a few themes before you add content.
I found that some of the theme did not play well with blogdown on GitHub.
And after I tried a few themes I think I accumulated a little baggage.
For example, I had directories names 'font' and 'fonts'.
I ended up deleting everything and starting again when I settled on a theme.


Once you have the template up and running you'll want to familiarize yourself with the blogdown documenation.
It can be found here:


```{r, eval=FALSE}
library(help = "blogdown")
```


You can use

```{r, eval=FALSE}
new_post(title="My new post", rmd = TRUE)
```


To create a template for a new post.
You can build it and view it locally as follows.


```{r, eval=FALSE}
serve_site()
```


If you like what you see you can commit and push your changes on git to the 'sources' branch.
If you followed the directions I did you can now run the 'deploy.sh' script and it should post to your GitHub site.

Happy blogging!


### Update 2017-06-02

I had trouble figuring out how to modify the menu in the theme Mainroad.
This may have to do with my inexperience with Hugo.
So I changed to the theme Hugo Bootstrag v4 Blog.
It seems simpler and may match my skill set better.

